---
title: "world_happiness_project2"
output: html_document
date: "2023-10-26"
---
```{r include=FALSE}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(janitor)
```

```{r include=FALSE}
world_happiness2015 <- read.csv("~/Downloads/archive/WHR_2015.csv", sep = ",")

world_happiness2016 <- read.csv("~/Downloads/archive/WHR_2016.csv", sep = ",")

world_happiness2017 <- read.csv("~/Downloads/archive/WHR_2017.csv", sep = ",")

world_happiness2018 <- read.csv("~/Downloads/archive/WHR_2018.csv", sep = ",")

world_happiness2019 <- read.csv("~/Downloads/archive/WHR_2019.csv", sep = ",")

world_happiness2020 <- read.csv("~/Downloads/archive/WHR_2020.csv", sep = ",")

world_happiness2021 <- read.csv("~/Downloads/archive/WHR_2021.csv", sep = ",")

world_happiness2022 <- read.csv("~/Downloads/archive/WHR_2022.csv", sep = ",")

world_happiness2023 <- read.csv("~/Downloads/archive/WHR_2023.csv", sep = ",")
```

```{r include=FALSE}






world_happiness2015v1 <- world_happiness2015%>%
  mutate(year = 2015)

world_happiness2016v1 <- world_happiness2016%>%
  mutate(year = 2016)

world_happiness2017v1 <- world_happiness2017%>%
  mutate(year = 2017)

world_happiness2018v1 <- world_happiness2018%>%
  mutate(year = 2018)

world_happiness2019v1 <- world_happiness2019%>%
  mutate(year = 2019)

world_happiness2020v1 <- world_happiness2020%>%
  mutate(year = 2020)

world_happiness2021v1 <- world_happiness2021%>%
  mutate(year = 2021)

world_happiness2022v1 <- world_happiness2022%>%
  mutate(year = 2022)

world_happiness2023v1 <- world_happiness2023%>%
  mutate(year = 2023)

world_happiness <- union( world_happiness2015v1, world_happiness2016v1)

world_happinessv1 <- union(world_happiness, world_happiness2017v1)

world_happiness2018v1$perceptions_of_corruption = as.numeric(as.character(world_happiness2018v1$perceptions_of_corruption))

world_happinessv2 <- union(world_happinessv1, world_happiness2018v1)

world_happinessv3 <- union(world_happinessv2, world_happiness2019v1)

world_happinessv4 <- union(world_happinessv3, world_happiness2020v1)

world_happinessv5 <- union(world_happinessv4, world_happiness2021v1)

world_happinessv6 <- union(world_happinessv5, world_happiness2022v1)

world_happinessv7 <- union(world_happinessv6, world_happiness2023v1)

world_happiness2018v1$perceptions_of_corruption = as.numeric(as.character(world_happiness2018v1$perceptions_of_corruption))


world_happinessv8 <- world_happinessv7%>%
  mutate(healthy_life_expectancy = coalesce(healthy_life_expectancy, 0.5840), perceptions_of_corruption = coalesce(perceptions_of_corruption, 0.13227))

world_happinessv8$country[world_happinessv8$country == 'Congo'] <- 'Congo (Brazzaville)'

world_happinessv8$country[world_happinessv8$country == 'Turkey'] <- 'Turkiye'

avg_hap_countby <- world_happinessv8%>%
  group_by(year) %>% 
  summarize(avg_happiness_score = mean(happiness_score))

avg_hap_countbyviz <- ggplot(avg_hap_countby, aes(year, avg_happiness_score))+
  geom_line(stat = "identity")+
  theme_classic()+
  scale_x_continuous(breaks = seq(2015, 2023, by = 1))+
  labs(title = "Overall Average Happiness Score", subtitle = "by year")

top10_hap_ey <- world_happinessv8 %>% 
  group_by(year) %>% 
  slice_max(happiness_score, n = 10)

top10_hap_eyviz <- ggplot(top10_hap_ey, aes(happiness_score, fill = country))+
  geom_histogram(bins = 30)+
  theme_classic()+
  labs(title = "Distribution of Top Ten Happiness from Each Year")


library(corrr)

happiness_score_corr<- world_happinessv8 %>% 
  select(happiness_score, gdp_per_capita, social_support, healthy_life_expectancy, freedom_to_make_life_choices, generosity, perceptions_of_corruption) %>% 
  correlate()

top50happinesscoreby <- world_happinessv8 %>% 
  group_by(year) %>% 
  slice_max(happiness_score, n = 50)

happiness_score_corrt50 <- top50happinesscoreby %>% 
  select(happiness_score, gdp_per_capita, social_support, healthy_life_expectancy, freedom_to_make_life_choices, generosity, perceptions_of_corruption) %>% 
correlate()

bottom50happinessscoreby <- world_happinessv8 %>% 
  group_by(year) %>% 
  slice_min(happiness_score, n = 50)

happiness_score_corrb50 <- bottom50happinessscoreby %>% 
  select(happiness_score, gdp_per_capita, social_support, healthy_life_expectancy, freedom_to_make_life_choices, generosity, perceptions_of_corruption) %>% 
  correlate()

overallavghappinessbc <- world_happinessv8 %>% 
  group_by(country) %>% 
  summarize(average_happiness_score = mean(happiness_score))
  
  topaveragehappinessbc <- overallavghappinessbc %>% 
    slice_max(average_happiness_score, n = 10)
  
  bottom10happinessbc <- overallavghappinessbc %>% 
  slice_min(average_happiness_score, n = 10)
  
  worldhappinessbr <- world_happinessv8 %>% 
  group_by(region) %>% 
  summarize(average_happiness_score = mean(happiness_score))
  
  world_happinessbrdesc <- worldhappinessbr %>% 
  arrange(desc(average_happiness_score))
  
  world_happinessbrbar <- ggplot(worldhappinessbr, aes(region, average_happiness_score, fill = region))+
  geom_bar(stat = "identity")+
  theme_excel()+
  labs(title = "Happiness Score by Region")+
  scale_y_continuous(breaks = seq(1, 8, by = 1))+
  theme(axis.text.x = element_blank())
  
  world_happinessv8$country[world_happinessv8$country == 'Somaliland region'] <- 'Somaliland Region'

```


After analyzing the happiness score in relation to some of the factors, I am curious how the countries compare by the  factors individually. The first column I am going to compare is gdp per captita. 


```{r}
gdppercapitadesc <- world_happinessv8 %>% 
  group_by(country) %>% 
  summarize(average_gdp = mean(gdp_per_capita)) %>% 
  arrange(desc(average_gdp))

print(gdppercapitadesc)

gdppercapitadesct12 <- gdppercapitadesc %>% 
  slice_max(average_gdp, n = 12)

print(gdppercapitadesct12)
```

Now I can visualize with ggplot...

```{r}
topaveragegdpviz <- ggplot(gdppercapitadesct12, aes(country, average_gdp, fill = country))+
  geom_bar(stat = "identity")+
  theme_classic()+
  scale_x_discrete(guide = guide_axis(n.dodge = 2))+
  labs(title = "Top 12 Average Gdp-Per-Capita", subtitle = "by country")+
  theme(axis.text.x = element_blank())

print(topaveragegdpviz)
```

Now I'll do the same for the social support column...

```{r}
socialsupportdesc <- world_happinessv8 %>% 
  group_by(country) %>% 
  summarize(average_social_support = mean(social_support)) 
  
socialsupportt12 <- socialsupportdesc %>% 
  slice_max(average_social_support, n = 12)

print(socialsupportt12)
  
```
Now I will plot it...

```{r}
socialsupportt12viz <- ggplot(socialsupportt12, aes(country, average_social_support))+
  geom_bar(stat = "identity", fill = "red")+
  theme_classic()+
  scale_x_discrete(guide = guide_axis(n.dodge = 2))+
  labs(title = "Top 12 Average Social Support", subtitle = "by country")

print(socialsupportt12viz)
```

From the ggpot visualize we can observe that Iceland has the highest social support score. The next column I am going to do is healthy life expectancy...

```{r}
lifeexpectancydesc <- world_happinessv8 %>% 
  group_by(country) %>% 
  summarize(average_life_expectancy = mean(healthy_life_expectancy)) %>% 
  arrange(desc(average_life_expectancy))

lifeexpectancyt12 <- lifeexpectancydesc %>% 
  slice_max(average_life_expectancy, n = 12)

print(lifeexpectancyt12)
```

Now the ggplot...

```{r}
lifeexpectancyt12viz <- ggplot(lifeexpectancyt12, aes(country, average_life_expectancy, fill = country))+
  geom_bar(stat = "identity")+
  theme_classic()+
  theme(axis.text.x = element_blank())+
  labs(title = "Top 12 Average Healthy Life Expectancy", subtitle = "by country")

print(lifeexpectancyt12viz)  
```
When looking at the visualization, it's clear that Hong Kong and Singapore scores highest with the average healthy life expectancies. The next column I will do is the freedom to make life choices column.

```{r}
freedomlifechoicesdesc <- world_happinessv8 %>% 
  group_by(country) %>% 
  summarize(average_freedom_to_make_life_choices = mean(freedom_to_make_life_choices)) %>% 
  arrange(desc(average_freedom_to_make_life_choices))

freedomlifechoicest12 <- freedomlifechoicesdesc %>% 
  slice_max(average_freedom_to_make_life_choices, n = 12)

print(freedomlifechoicest12)



```
Now to visualize...

```{r}
freedomlifechoicest12viz <- ggplot(freedomlifechoicest12, aes(country, average_freedom_to_make_life_choices))+
  geom_bar(stat = "identity", fill = "green")+
  theme_classic()+
  scale_x_discrete(guide = guide_axis(n.dodge = 2))+
  labs(title = "Top 12 Average Freedom to Make Life Choice", subtitle = "by country")

print(freedomlifechoicest12viz)
```
From the visualization, Uzbekistan leads in the freedom to make life choices score. The next score I am going to look at is the generosity score. I will do as I did with the previous columns.

```{r}
generositydesc <- world_happinessv8 %>% 
  group_by(country) %>% 
  summarize(average_generosity = mean(generosity)) %>% 
  arrange(desc(average_generosity))

generosityt12 <- generositydesc %>% 
  slice_max(average_generosity, n = 12)

print(generosityt12)
```
After arranging the average generosity in descending order and taking the top 12, Myanmar is the has the highest generosity score. This of course be better visualized with ggplot...

```{r}
generosityt12viz <- ggplot(generosityt12, aes(country, average_generosity))+
  geom_bar(stat = "identity", fill = "brown")+
  scale_x_discrete(guide = guide_axis(n.dodge = 2))+
  theme_classic()+
  labs(title = "Top 12 Average Generosity Score", subtitle = "by country")

print(generosityt12viz)
```

As you can see from the bar chart it clearly show's Myanmar leading in the generosity category.
Now I am going to check the last variable of this dataset, perceptions of corruption...

```{r}
perceptionsofcorruptiondesc <- world_happinessv8 %>% 
  group_by(country) %>% 
  summarize(average_perceptions_of_corruption = mean(perceptions_of_corruption)) %>% 
  arrange(desc(average_perceptions_of_corruption))

perceptionsofcorruptiont12 <- perceptionsofcorruptiondesc %>% 
  slice_max(average_perceptions_of_corruption, n = 12)

print(perceptionsofcorruptiont12)
```
After running that query, Singapore has the highest score in this category. Like I did with the other categories, I will now visualize this.

```{r}
perceptionsofcorruptiont12viz <- ggplot(perceptionsofcorruptiont12, aes(country, average_perceptions_of_corruption))+
  geom_bar(stat = "identity", fill = "chartreuse4")+
  scale_x_discrete(guide = guide_axis(n.dodge = 2))+
  theme_classic()+
  labs(title = "Top 12 Average 'Perceptions of Corruption' Scores", subtitle = "by country")

print(perceptionsofcorruptiont12viz)


```

As the chart clearly shows, Singapore leads in this category. 

Well, that concludes my world_happiness project. Over the course of this project I cleaned and fixed several errors to make sure the data was ready for analysis. I then began analyzing where I found out many things, such as gdp per capita is the variable that positively correlates with the happiness score for dataset as a whole. Thanks for joining me, I hope to only improve as I continue these projects.